Tropical Mathematics and the Lambda-calculus I

### 2023-11-30 CSL24/section2bisbis.tex
commit

This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}.
Typical questions in this case would be, for a fixed $\omega_0$:
%
%The tropical point of view allows now to express two natural questions about this situation:
\begin{enumerate}
 \item What is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0} \false$''  (similarly for $\true$)?
 I.e., which is the choice of $p,q$ that maximizes the probability $P_{\omega_0}$?
 \item 
Knowing that $M$ reached a normal form, %produced $\false$ (similarly for $\true$),
 which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0}\false$''? %, knowing that ``$M\twoheadrightarrow \false$''?
I.e., which is the choice of $p,q$ that makes $\omega_0$ the most likely path %among those leading to $\false$ 
(i.e.\ that maximizes the conditional probability $\BB P(``M\twoheadrightarrow_{\omega_0} \false'')$?% \mid ``M\twoheadrightarrow \false'')$)?
\end{enumerate}

Answering 1) amounts at solving a contrained maximization problem:
\[(p,q)\in\arg\max_{y=1-x} P_{\omega_{0}}(x,y)\] or, equivalently: $p\in\arg\max_{x\in[0,1]} P_{\omega_0}(x,1-x)$,
related to the monomial $P_{\omega_{0}}(p,q)$ and for which a tropical transformation does not, in general, help factoring the contraint $q=1-p$ out of the maximization.
We will, instead, mainly concentrate on 2).
This asks for finding a solution to the following contrained maximization problem in the unknown $p\in[0,1]$:

@@

P_{\omega_0}(p,1-p) = \max_w P_{\omega}(p,1-p)

@@

Now, the difficulty raised by the constraint between the two variables of the monomials might be overcome by passing to the associated \emph{tropical} polynomial.
In fact, since $-\log P_\omega(p,q)=(\trop P)(-\log p,-\log(1-p))$, this is equivalent to the following minimization problem in the unknown $p\in[0,1]$:
(\trop P_{\omega_0})(-\log p,-\log(1-p)) = (\trop Q)(-\log p,-\log(1-p))
and with a change of variable we can now move the difficult constraint outside the minimization problem, as this is equivalent to finding a solution of the following system of equations in the unknowns $(p,x,y)\in[0,1]\times[0,\infty]\times[0,\infty]$:
\trop P_{\omega_0}(x,y) = \trop Q (x,y), \qquad x= -\log p,\qquad y=-\log(1-p).
Since studying a tropical polynomial could be a feasable task, the first equation might be easily trasformed into an explicit relation between $x$ and $y$, and this might make it possible to solve the whole system, finally finding our $p$, as the next example shows.

@@

For our running example $M$, we have $\trop Q_{\true}(x,y)=\min\set{2x,y+2x,3y}$.
For simplicity, let us suppose that we observed the event ``$M\twoheadrightarrow\true$'', and thus our probabilities are conditioned under this observation.
It is not difficult to see that $\trop Q_{\true}(x,y)=3y$ iff $y\leq \frac{2}{3}x$, and $\trop Q_{\true}(x,y)=2x$ otherwise.
Remembering that $3y=\trop P_{rrr}(x,y)$, we can now solve problem 2) for $\omega_0=rrr$:

@@

This means that, if $p\in[0,1]$ satisfies $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), then $P_{rrr}(p,1-p) = \max_{\omega=ll, rll, rrr} P_{\omega}(p,1-p)$, i.e.: knowing that $M$ sampled $\true$ in its normal form, the most likely sampled occurrence of $\true$ is the one at the address $rrr$ (the rightmost one) iff $1-p\geq p^{\frac{2}{3}}$.
Said differently, the maximum likelihood estimator of $p$, with prior knowledge  ``$M\twoheadrightarrow \true$'' and seen the event ``$M\twoheadrightarrow_{\omega_0}\true$'', is any $p$ s.t.\ $1-p\geq p^{\frac{2}{3}}$. We have thus solved problem 2) above for $\omega_0=rrr$, under the additional hypothesis that ``$M\twoheadrightarrow\true$''.
Remembering that $2x=\trop P_{ll}(x,y)$, in this simple case we can even say that for the other values of $p$ (for example, $p=\frac{1}{2}$), the most likely $\true$ to be sampled (knowing that a $\true$ has been sampled) is the one at address $ll$, therefore actually solving problem 2), under the same additional hypothesis, for all instances of the reduction path $\omega$ leading to $\true$ (the path $rll$ being the most likely one for no $p$ at all).

@@

As we'll see in Section \ref{section3}, this analysis extends to PCF-style programs. For example, the program $M=\mathbf Y(\lambda x.\true \oplus_{p} x)$ yields the power series $Q_{\true}(p,q)=\sum_{n=0}^{\infty}pq^{n}=\frac{p}{1-q}$ that sums all \emph{infinitely many} ways in which $M$ may reduce to $\true$. Notice that the tropicalised series $\mathsf tQ_{\true}(-\log p,-\log q)=\inf_{n\in \mathbb N}\{-\log p -n\log q\}=-\log p$ collapses onto a single monomial describing the \emph{unique} most likely reduction path of $M$, namely the one that passes through a coin toss only once. 
%
%Similar arguments could be done by replacing $\true$ and $\false$ by, respectively, a converging and diverging term (e.g.~in a $\mathrm{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.

@@

This example is inspired from \cite{Manzo2013}. We consider now a basic non-deterministic language with terms $M::= \true \mid  \mathtt{Gen} \mid  M + M$, with an operation semantics comprising a non-deterministic reduction rule 
$M_{1}+M_{2} \stackrel{\alpha}{\to} M_{i}$ and a generation rule
$\mathtt{Gen}\stackrel{\beta}{\to} \true +\mathtt{Gen}$, 
where in each case the value $\alpha,\beta\in \Lawv$ indicates a \emph{cost} associated with the reduction (e.g.~the estimated clock value for the simulation of each reduction on a given machine model). 
Then, any reduction $\omega: M \twoheadrightarrow N$ of a term to (one of its) normal form is  associated with a tropical monomial $P_{\omega}( \alpha,\beta)$ consisting of the sum of the costs of all reductions in $\omega$. For a given normal form $N$, the reductions $\omega_{i}: M \twoheadrightarrow N$ give rise to a tps $\inf_{i\in I}P_{\omega_{i}}( \alpha,\beta)$. 
For example, consider the non-deterministic term
$
M :=\mathtt{Gen}+  ((\true + \true) + \mathtt{Gen})
$. 
The (infinitely many) reduction paths leading to $\true$ can be grouped as follows:
\begin{itemize}
\item left, then reduce $\mathtt{Gen}$ $n+1$-times, then left;
\item right, then left and then either left or right;
\item right twice, then reduce $\mathtt{Gen}$ $n+1$-times and then left.
\end{itemize}
This leads to the tps 
$\varphi_{M\twoheadrightarrow \true}(\alpha)=\inf_{n\in \mathbb N}\big\{2\alpha+(n+1)\beta, 3\alpha, 3\alpha+(n+1)\beta\big \}= \min\{ 2\alpha+\beta, 3\alpha\}$, which describes all possible behaviors of $M$. Notice that, since $\alpha$ and $\beta$ are always positive, the power series $\varphi_{M\twoheadrightarrow\true}(\alpha)$ is indeed equivalent to the tropical polynomial $\min\{ 2\alpha+\beta, 3\alpha\}$. In other words, of the infinitely many behaviors of $M$, only finitely many have chances to be \emph{optimal}: either left + $\mathtt{Gen}$ + left, or right + left + (left,right). Also in this case, reducing to best-case analysis leads to collapse the infinitary description of \emph{all} behaviors to a purely combinatorial description of the finitely many optimal ones.

>>>>

This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}.
A natural question in this case would be, for a fixed $\omega_0$:
%
%The tropical point of view allows now to express two natural questions about this situation:
%\begin{enumerate}
% \item What is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0} \false$''  (similarly for $\true$)?
% I.e., which is the choice of $p,q$ that maximizes the probability $P_{\omega_0}$?
% \item 
knowing that $M$ reached a normal form, say $\true$, %produced $\false$ (similarly for $\true$),
 what is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow_{\omega_0}\true$''? %, knowing that ``$M\twoheadrightarrow \false$''?
In other words, what is the choice of $p,q$ that maximizes the conditional probability $\BB P(M\twoheadrightarrow_{\omega_0} \true\mid M\twoheadrightarrow \true)$, i.e.~that 
 makes $\omega_0$ the most likely path among those leading to $\true$?
% \mid ``M\twoheadrightarrow \false'')$)?
%\end{enumerate}
%
%Answering 1) amounts at solving a constrained maximization problem:
%\[(p,q)\in\arg\max_{y=1-x} P_{\omega_{0}}(x,y)\] or, equivalently: $p\in\arg\max_{x\in[0,1]} P_{\omega_0}(x,1-x)$,
%related to the monomial $P_{\omega_{0}}(p,q)$ and for which a tropical transformation does not, in general, help factoring the contraint $q=1-p$ out of the maximization.
%We will, instead, mainly concentrate on 2).

Answering this question amounts at finding a solution to the following constrained maximization problem in the unknown $p\in[0,1]$:
%\[
%P_{\omega_0}(p,q) = \max_\omega P_{\omega}(p,q)
%
%
%\]
%
%\[
%(p,q) \in \arg\max_{y=1-x} P_{\omega_{0}}(x,y)\] or, equivalently: $p\in\arg\max_{x\in[0,1]} P_{\omega_0}(x,1-x)$,
%
%\]
% in the unknown $p\in[0,1]$:

@@

P_{\omega_0}(p,1-p) = \max_\omega P_{\omega}(p,1-p)

@@

% the difficulty raised by the constraint between the two variables of the monomials might be overcome by passing to the associated \emph{tropical} polynomial.
%In fact, 
Since $-\log P_\omega(p,q)=(\trop P_{\omega})(-\log p,-\log q)$, this is equivalent
% to the following \emph{minimization} problem in the unknown $p\in[0,1]$:
%\[
%(\trop P_{\omega_0})(-\log p,-\log(1-p)) = (\trop Q_{\true})(-\log p,-\log(1-p))
%\]
%%With a change of variable we can now move the  constraint outside the minimization problem, as
%which is in turn equivalent
 to finding a solution of the following constrained \emph{minimization} problem in the unknowns $(p,x,y)\in[0,1]\times[0,\infty]\times[0,\infty]$:
\begin{equation}\label{eq:troproot}
\trop P_{\omega_{0}} (x,y)= \trop Q_{\true}(x,y) , \qquad x= -\log p,\qquad y=-\log(1-p).
\end{equation}
Since the first equation can be solved easily (i.e.~in linear time) by computing the tropical roots of $\trop Q_{\true}$, we can obtain an explicit relation between $x$ and $y$ that can be used to solve the whole system, finally finding our $p$, as the next example shows.

@@

For our running example $M$, 
let us suppose that we observed the event ``$M\twoheadrightarrow\true$'', and thus our probabilities are conditioned under this observation. We have
\[\trop Q_{\true}(x,y)=\min\set{\trop P_{ll}(x,y),\trop  P_{rll}(x,y),\trop P_{rrr}(x,y)}=\min
\set{2x,y+2x,3y}.\]
The tropical roots of $\trop Q_{\true}(x,y)$ are $(0,0)$ and all points of the form $(x,\frac{2}{3}x)$. Recall that these are the points where \autoref{eq:troproot} is satisfied for \emph{at least two} distinct values of $\omega_{0}$. From this it follows that $\trop Q_{\true} (x,y)=\trop P_{rrr}(x,y)=3y$ holds iff $y\leq \frac{2}{3}x$, and  $\trop Q_{\true}(x,y)=2x=\trop P_{ll}(x,y)$ otherwise.
In this way can find the maximum likelihood estimator for $\omega_0=rrr$:

@@

This means that, if $p\in[0,1]$ satisfies $1-p\geq p^{\frac{2}{3}}$ (for example, $p=\frac{1}{4}$), then $P_{rrr}(p,1-p) = \max_{\omega=ll, rll, rrr} P_{\omega}(p,1-p)$, i.e.: knowing that $M$ sampled $\true$ in its normal form, the most likely sampled occurrence of $\true$ is the one at the address $rrr$ iff $1-p\geq p^{\frac{2}{3}}$.
%Said differently, the maximum likelihood estimator of $p$, with prior knowledge  ``$M\twoheadrightarrow \true$'' and seen the event ``$M\twoheadrightarrow_{\omega_0}\true$'', is any $p$ s.t.\ $1-p\geq p^{\frac{2}{3}}$. 
%Observe that the tropical roots of $\trop Q_{\true}(x,y)$ provide a complete description of the relationship between events and their maximum likelihood estimators. 
%We have thus solved problem 2) above for $\omega_0=rrr$, under the additional hypothesis that ``$M\twoheadrightarrow\true$''.
%Remembering that $2x=\trop P_{ll}(x,y)$, in this simple case we can even say that for the other values of $p$ (for example, $p=\frac{1}{2}$), the most likely $\true$ to be sampled (knowing that a $\true$ has been sampled) is the one at address $ll$, therefore actually solving problem 2), under the same additional hypothesis, for all instances of the reduction path $\omega$ leading to $\true$ (the path $rll$ being the most likely one for no $p$ at all).

@@

As we'll see in Section \ref{section3}, this analysis extends to PCF-style programs. For example, the program $M=\mathbf Y(\lambda x.\true \oplus_{p} x)$ yields the power series $Q_{\true}(p,q)=\sum_{n=0}^{\infty}pq^{n}=\frac{p}{1-q}$ that sums all \emph{infinitely many} ways in which $M$ may reduce to $\true$. Notice that the tropicalised series $\mathsf tQ_{\true}(-\log p,-\log q)=\inf_{n\in \mathbb N}\{-\log p -n\log q\}=-\log p$ collapses onto a single monomial describing the \emph{unique} most likely reduction path of $M$ leading to $\true$, namely the one that passes through a coin toss only once. 
%
%Similar arguments could be done by replacing $\true$ and $\false$ by, respectively, a converging and diverging term (e.g.~in a $\mathrm{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.

@@

This example is inspired from \cite{Manzo2013}. We consider now a basic non-deterministic language with terms $M::= \true \mid  \mathtt{Gen} \mid  M + M$, with an operation semantics comprising a non-deterministic reduction rule 
$M_{1}+M_{2} \stackrel{\alpha}{\to} M_{i}$ and a generation rule
$\mathtt{Gen}\stackrel{\beta}{\to} \true +\mathtt{Gen}$, 
where in each case the value $\alpha,\beta\in \Lawv$ indicates a \emph{cost} associated with the reduction (e.g.~the estimated clock value for the simulation of each reduction on a given machine model). 
Then, any reduction $\omega: M \twoheadrightarrow N$ of a term to (one of its) normal form is  associated with a tropical monomial $P_{\omega}( \alpha,\beta)$ consisting of the sum of the costs of all reductions in $\omega$. For a given normal form $N$, the reductions $\omega_{i}: M \twoheadrightarrow N$ give rise to a tps $\inf_{i\in I}P_{\omega_{i}}( \alpha,\beta)$. 
For example, consider the non-deterministic term
$
M :=\mathtt{Gen}+  ((\true + \true) + \mathtt{Gen})
$. 
The (infinitely many) reduction paths leading to $\true$ can be grouped as follows:
\begin{itemize}
\item left, then reduce $\mathtt{Gen}$ $n+1$-times, then left;
\item right, then left and then either left or right;
\item right twice, then reduce $\mathtt{Gen}$ $n+1$-times and then left.
\end{itemize}
This leads to the tps 
$\varphi_{M\twoheadrightarrow \true}(\alpha)=\inf_{n\in \mathbb N}\big\{(n+2)\alpha+(n+1)\beta, 3\alpha, (n+3)\alpha+(n+1)\beta\big \}= \min\{ 2\alpha+\beta, 3\alpha\}$, which describes all possible behaviors of $M$. Notice that, since $\alpha$ and $\beta$ are always positive, the power series $\varphi_{M\twoheadrightarrow\true}(\alpha)$ is indeed equivalent to the tropical polynomial $\min\{ 2\alpha+\beta, 3\alpha\}$. In other words, of the infinitely many behaviors of $M$, only finitely many have chances to be \emph{optimal}: either left + $\mathtt{Gen}$ + left, or right + left + (left,right). Also in this case, reducing to best-case analysis leads to collapse the infinitary description of \emph{all} behaviors to a purely combinatorial description of the finitely many optimal ones.

### 2023-11-30 CSL24/section3fancy.tex
commit

Now, one can see that the comonad $!$ of $\LREL$ can be ``decomposed'' into a family of ``graded exponentials functors'' $!_n:\LREL\to\LREL$ ($n\in\BB N$), where $!_{n}X$ is the set of multisets on $X$ of cardinality \emph{at most} $n$. %  lift to functors 
The sequence $(!_n)_{n\in\N}$ gives rise to a so-called \emph{$\N$-graded linear exponential comonad} on (the SMC) $\LREL$ \cite{Katsumata2018}. %satisfying the adjunction: $\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
As such, $(\LREL,(!_n)_{n\in\N})$ yields then a model of $\BSTLC$. Remark that arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$. Notice that, whenever $\model *$ is finite, the set $\model{A}$ is finite for any type $A$ of $\BSTLC$.

@@

Applying \cite[Theorem 6.1]{lemay2020} one can check that
%\begin{proposition}[{\color{red}LEMAY??}]\label{thm:LREL!CCDC}
 $\LREL_!$ is a CC$\partial\lambda$C (see Section \ref{section5bis}) when equipped with $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the iso $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
 The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tps $f:\Lawv^{X}\to \Lawv^{Y}$ into a tps $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by 
$D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a}+x_{a}+\mu y\right\}$. One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
%Moreover, the Taylor formula \eqref{eq:taylorcat} yields a ``tropical'' Taylor formula for tps of the form 
%$f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}$. 
%\end{proposition}
%\begin{remark}For $t\in\HOM{\LREL}{!X}{Y}$, we have: $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu+\nu'+\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho+\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.\end{remark}
%This ensures that one can define a sound interpretation of $\STDLC$-terms in the standard way (see [Section 4.3, \cite{Manzo2010}]).
%As such, $(\LREL_!,D)$ is a model of the differential $\lambda$-calculus $\STDLC$ \cite{ER}.

>>>>

On the side of $\LREL$, one can check that the comonad $!$ of $\LREL$ can be ``decomposed'' into a family of ``graded exponentials functors'' $!_n:\LREL\to\LREL$ ($n\in\BB N$), where $!_{n}X$ is the set of multisets on $X$ of cardinality \emph{at most} $n$. %  lift to functors 
The sequence $(!_n)_{n\in\N}$ gives rise to a so-called \emph{$\N$-graded linear exponential comonad} on (the SMC) $\LREL$ \cite{Katsumata2018}. %satisfying the adjunction: $\LREL(Z\otimes !_{n}X,Y) \simeq \LREL(Z, !_{n}X\multimap Y)$.
As such, $(\LREL,(!_n)_{n\in\N})$ yields then a model of $\BSTLC$. More details can be found in the extended version of this article.

Notice that arrow types are interpreted via $\model{!_{n}A\multimap B}:= !_{n}\model A \times \model B$ and that, whenever $\model *$ is finite, the set $\model{A}$ is finite for any type $A$ of $\BSTLC$.

@@

Applying \cite[Theorem 6.1]{lemay2020} one can check that
%\begin{proposition}[{\color{red}LEMAY??}]\label{thm:LREL!CCDC}
 $\LREL_!$ is a CC$\partial\lambda$C when equipped with $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X\& X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (using the isomorphism $(\mu,\rho)\in !Z\times !Z'\mapsto\mu\oplus\rho \in !(Z+Z')$).
 The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tps $f:\Lawv^{X}\to \Lawv^{Y}$ into a tps $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by 
$D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a,b}+x_{a}+\mu y\right\}$. %, where $\hat f \in \LREL^{!X\times Y}$ is the matrix associated with $f$.
One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
%Moreover, the Taylor formula \eqref{eq:taylorcat} yields a ``tropical'' Taylor formula for tps of the form 
%$f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}$. 
%\end{proposition}
%\begin{remark}For $t\in\HOM{\LREL}{!X}{Y}$, we have: $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu+\nu'+\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho+\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.\end{remark}
%This ensures that one can define a sound interpretation of $\STDLC$-terms in the standard way (see [Section 4.3, \cite{Manzo2010}]).
%As such, $(\LREL_!,D)$ is a model of the differential $\lambda$-calculus $\STDLC$ \cite{ER}.
